---
title: "Assignment"
author: "Me"
date: "29/12/2020"
output: html_document
---

**Reading the data into R**
```{r echo=TRUE, cache=TRUE}
data<- read.csv('activity.csv')
head(data)
```

Now we are preprocessing the data. 

First we are changing the class of the date variable
into date class using the lubridate package.

```{r echo=TRUE}
library(lubridate)
data$date<- ymd(data$date) #converting the date variable into date format
str(data)
head(data)
```

Finding sum of steps per day and turning it into a dataframe

```{r echo=TRUE}
x<-as.data.frame(tapply(data$steps, data$date, sum)) 
names(x)<- 'total_steps_per_day'
x$Date<- rownames(x)
```

Now plotting the histogram for total no. of steps per day

```{r echo=TRUE}
hist(x = x$total_steps_per_day, xlab = 'Total no. of steps/day', main = 'Histogram for total no. of steps per day', col = x$total_steps_per_day )
```

Now showing the mean and median of total no. of steps/day

```{r echo=TRUE}
summary(x$total_steps_per_day)
```

*Finding average daily activity pattern and showing it in a plot*

Lets first remove the NA values from the data and keep it in a fresh dataframe

```{r echo=TRUE}

data1<- data[complete.cases(data),] #removing NA values from the dataset and assigning it to new dataset 
head(data1)
```

First finding the average/mean values and putting it into a dataframe adp for convenience

```{r echo=TRUE}
adp<- as.data.frame(tapply(data1$steps, data1$interval, mean))
names(adp)<- 'Avg.daily.steps'
adp$intervals<- rownames(adp)
head(adp)
```

Finding the 5 min interval which has the most no. of steps on average

```{r echo=TRUE}
max(adp$Avg.daily.steps) #finding the max of Avg.daily.steps
```

So 206.1698 is the maximum value of Avg.daily.steps. So now finding the interval containing the max value of Avg.daily.steps i.e. 206.1698

```{r echo=TRUE}
adp[grep(206.1698, adp$Avg.daily.steps),]
```

now creating the graph

```{r echo=TRUE}
plot(adp$intervals, adp$Avg.daily.steps, xlab = '5 minute interval', ylab = 'average number of steps taken', type = 'l', main = 'Average daily pattern', sub = 'interval 835 on average across all days has the most number of steps')
abline(h = max(adp$Avg.daily.steps))
```

*Imputing missing values and creating a new dataset*
Finding number of missing values in the original dataset

```{r echo=TRUE}
miss_value<- sum(is.na(data$steps))
```

Filling missing values with the average/mean value of total no. of steps of that particular interval

```{r echo=TRUE}
data2<- data #copying the orginal dataframe 'data' into a new dataframe called 'data2'
#imputing missing or NA values with the mean no. of steps for that particular interval
for (i in 1:nrow(data2)){
  if(is.na(data2$steps[i])==TRUE){ #if missing value is found then the imputation is #done
    a<- data2$interval[i]
    b<- adp[adp$intervals==a,][1,1]
    data2$steps[i]<-b #replacing missing NA value with mean value
    a<-0 #initializing a= 0 for next iteration of the loop
    b<-0 #initializing b= 0 for next iteration of the loop
  }
}
head(data2) #now data2 has all NA values replaced by the mean of total no. of steps for that particular interval
```

**Now making comparison of total no. of steps per day between the old and new datasets i.e. data and data2**
First finding the total no. of steps taken and keeping it in a dataframe y

```{r echo=TRUE}
y<- as.data.frame(tapply(data2$steps, data2$date, sum)) 
names(y)<- 'total_steps_per_day'
y$Date<- rownames(y)
```

Showing summary stats i.e. mean and median for total no. of steps from the new dataset
```{r echo=TRUE}
summary(y$total_steps_per_day)
```

Showing summary stats i.e. mean and median for total no. of steps from the original dataset

```{r echo=TRUE}
summary(x$total_steps_per_day)
```

*so there is no change in mean and only a negligible impact on median. However the 1st and 3rd quartile values have changed*

Now making 2 histograms in 2 panels showing differences in the total no. of steps for the old and new datasets

```{r echo=TRUE}
par(mfrow= c(1,2))

hist(x = x$total_steps_per_day, xlab = 'Total no. of steps/day', main = 'original data with missing values', col = x$total_steps_per_day )
hist(x= y$total_steps_per_day, xlab = 'Total no. of steps/day', main = 'new dataset with imputed missing values', col = y$total_steps_per_day)
```

Now checking for differences in activity patterns between weekdays and weekends using data2

```{r echo=TRUE}
data2$weekday<- weekdays(data2$date) #finding the weekdays 
#now creating a factor variable that will show whether its a weekday or weekend
data2$weekday_factor<- data2$weekday %in% c('Monday','Tuesday', 'Wednesday','Thursday','Friday')
data2$weekday_factor<- as.factor(data2$weekday_factor)
```

Now separating this dataset into weekdays and weekends

```{r echo=TRUE}
weekdays<- data2[data2$weekday_factor==TRUE,]
weekends<- data2[data2$weekday_factor==FALSE,]
```

Now finding average no. of steps for all 5 min intervals for weekdays and keeping it in a separate data frame for convenience.

```{r echo=TRUE}
weekdays_average<-  as.data.frame(tapply(weekdays$steps, weekdays$interval, mean))
names(weekdays_average)<- 'Avg.daily.steps'
weekdays_average$interval<- rownames(weekdays_average)

```

Now finding average no. of steps for all 5 min intervals for weekends and keeping it in a separate data frame for convenience.

```{r echo=TRUE}
weekends_average<- as.data.frame(tapply(weekends$steps, weekends$interval, mean))
names(weekends_average)<- 'Avg.daily.steps'
weekends_average$interval<- rownames(weekends_average)
```

**now creating panel plot of 2 timeseries graphs of type 'l' for showing differences in average number of steps taken, averaged across all weekday days or weekend days (y-axis)**

```{r echo=TRUE}
par(mfrow= c(1,2))

plot(weekdays_average$interval, weekdays_average$Avg.daily.steps, type = 'l', xlab = '5 min intervals', ylab = 'average no. of steps', col= weekdays_average$Avg.daily.steps, main = 'weekdays')
plot(weekends_average$interval, weekends_average$Avg.daily.steps, type = 'l', xlab = '5 min intervals', ylab = 'average no. of steps', col= weekdays_average$Avg.daily.steps, main= 'weekends')
```










